Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-01-06 doc/invt.tex
small edits

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and metadynamics\cite{huber1994, laio2002,
laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
flat-distribution-sampling (FDS) techniques
that build the bias potential adaptively or on the fly.
%
Both techniques employ regular updating of the bias potential
to discourage future visits to the current configuration
by incrementally elevating the bias potential there.
%
The difference lies
in the updating hill function,
which takes the form of a localized
$\delta$-function in the WL case,
or that of a Gaussian
in the metadynamics.
%
%is that in the WL case,
%each updating step is limited to the bin
%containing the current $z$,
%while metadynamics adopts an extended
%Gaussian window
%covering several neighboring bins.
%%
%The latter is more often used in MD simulations
%as it guarantees a smooth profile
%of the bias potential
%suitable for numerical differentiation
%in deriving the bias force.
%
In both cases,
regular updates to the bias potential
%in both cases, the bias potential
%is updated frequently (usually every few MC or MD steps),
%and the fluctuation
disrupts the underlying equilibrium
sampling\cite{laio2002},
and for convergence
one  has to gradually decrease
the updating magnitude.

The schedule of reducing
the updating magnitude,
therefore, is critical in improving
the precision of the final bias potential,
hence that of the PMF\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014,
barducci2008, dickson2011, dama2014}.
%
For the WL algorithm, it is known
that the optimal schedule
is a function of the
inverse time\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008}
or
the inverse number of simulation steps.
%
The optimal schedule for the metadynamics case,
however, is less studied.

In this study,
we present a method of computing
the optimal schedule
for a general adaptive FDS method,
including the WL algorithm and metadynamics.
%
We derive the optimal schedule
from an equation of motion of a free particle,
and show that the initial updating magnitude
should be set to
half of the previous equilibrium value.
%
The optimal schedule in the WL case
is shown to be given by
an inverse-time formula,
with a modified origin of time.
%
However,
the optimal schedule for metadynamics is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
Further, we generalize
the WL updating scheme
to a class of bandpass updating schemes
that are optimal for long simulations,
and the above origin-modified inverse-time schedule
turns out to be optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.
%
These results may hopefully help one to better
plan and optimize FDS simulations.

@@

Consider the problem of computing
the distribution, $p^*_i$,
along a discrete quantity, $i$,
for a given system.
%
%The discreteness of $i$ is natural
%in discrete models.
%
For example, $i$ can be the energy, $E$,
in a lattice spin model or the temperature index
in a simulated tempering simulation\cite{
marinari1992, lyubartsev1992}.
%
For a continuous quantity, $z$,
which is typical in a molecular system,
we can discretize $z$
such that the integer $i$ represents
the index of a small interval, or a bin,
$(z, z + \Delta z)$.
%
In both cases,
the distribution is normalized as
$\sum_{i = 1}^n p^*_i = 1$.

@@

Here we shall call a simulation that targets
a nearly flat distribution\cite{
dayal2004, trebst2004, zhang2007, barducci2008, singh2011},
$p_i$,
an FDS simulation.
%
Roughly speaking, there are two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
uses a fixed bias potential, $V_i$,
(which is estimated beforehand)
during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{
berg1992, lee1993}
and simulated tempering\cite{
marinari1992, lyubartsev1992}
with $i$ being the index of energy
and that of temperature, respectively.
%
The estimated bias potentials at bin $i$, $V_i$,
are often improved
with the histogram of the sampling, $H_i$, accumulated
during simulation.
%
Using $H_i$ as an estimate of $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1}
allows us to correct $V_i$ as
%
\begin{equation}
V^\mathrm{corr}_i
=
V_i
+
\ln \frac{ H_i }
         { p_i }.
\label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    H_i \approx \pi_i
    &\propto p^*_i \, e^{-V_i},
    \\
    p_i
    &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
For an unknown system,
it is often difficult
to guess a sufficiently accurate
$V_i$ to start with.
%
Thus,
it can take many iterative rounds of simulations
to achieve the desired flat distribution.
%
\note{Further, a reasonably precise histogram
requires an accumulation period
longer than the autocorrelation time,
which can also be hard to estimate beforehand.}

For this reason, the second type of methods,
referred to as the adaptive FDS methods,
were invented.
%
Examples include
the WL algorithm and metadynamics.
%
These methods incrementally update the bias potential
and can approximately sample the target distribution.
%
However,
%they are less well understood theoretically, as
as mentioned, the updates  break the microscopic reversibility
of the underlying sampling mechanism. Further failure to properly reduce the size of the histogram increments would lead to roughness or errors in the final estimate.
%
%Thus, for a conservative practitioner,
%the E-FDS methods appear more appealing
%once a sufficiently accurate bias potential is available.
%
To minimize this effect,
many adaptive methods adopt a schedule to
reduce the updating magnitude
over the course of simulation\cite{
marsili2006,
liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
barducci2008}.
%
Thus, in late stages of a long simulation,
the updating becomes smaller and
%the adaptive simulation tends to
%the corresponding equilibrium one.
%
if the bias potential were converged to the exact value,
this adaptive FDS tends to the ideal equilibrium counterpart,
%particular equilibrium simulation can be thought as
%the optimal asymptotic limit of the adaptive counterpart,
delivering the maximal efficiency.

@@

One can show that with a constant $\alpha(t) = \alpha > 0$,
the distribution collected from
an adaptive FDS simulation
is identical to $p_i$ in long times.
%
However, as the above updates
change $v_i(t)$ continuously,
Eq. \eqref{eq:pi_p_v} no longer holds
at all times,
and the error in $v_i(t)$
depends on the updating magnitude\cite{
  zhou2005, liang2005, laio2005, bussi2006, poulain2006, liang2007,
  morozov2007, zhou2008, morozov2009, crespo2010,
  atchade2011, fort2015}.
%%
%The source of the deviation is two-fold.
%%
%First, since $v_i(t)$ is updated continuously,
%there is a random noise that is proportional
%to $\sqrt \alpha$.
%%
%Second, there is a systematic error
%that comes from the updating dynamics itself,
%as it breaks the Markovian nature
%that underlies Eq. \eqref{eq:pi_p_v}.
%%
%
Thus, one has to decrease $\alpha$ over time
to improve the estimate of the distribution,
$p^*_i$. %via Eq. \eqref{eq:v_def}.

@@

We first approximate the finite difference Eq. \eqref{eq:wl_update}
by a differential equation
%
\begin{equation}
  \dot v_i(t)
  =
  \alpha(t) \, \frac{ h_i(t) } { p_i },
  \label{eq:vt_diffeq}
\end{equation}
%
where
$\dot v_i(t) \equiv dv_i(t)/dt$,
%
and
\begin{equation}
  h_i(t) = \delta_{i, i(t)}
  ,
  \label{eq:h_def}
\end{equation}
is the instantaneous histogram,
which is equal to $1.0$
for the bin $i(t)$ at time $t$
or zero otherwise.

Next, we split $h_i(t)$ into a deterministic part
and a stochastic part,
%
\begin{equation}
  h_i(t) = \langle h_i(t) \rangle + \zeta_i(t).
  \label{eq:h_split}
\end{equation}
%
The deterministic part can be related
to the ``ensemble average'' of $h_i$,
where the ensemble consists of many similar simulation copies
sharing the same schedule $\alpha(t)$
the same $v_i(t)$ at time $t$.
%
The initial states and the stochastic forces
during the process may differ, however.

For sufficiently small $\alpha(t)$,
the $v_i$'s remain roughly the same for a short period.
%
Then,
the sampling process is approximately Markovian, % of a finite order,
and we may assume Eq. \eqref{eq:pi_p_v}
for the deterministic part
%
\begin{equation}
  \langle h_i(t) \rangle
  \approx
  \pi_i(t)
  =
  \frac{                p_i \, e^{-v_i(t)} }
       { \sum_{j = 1}^n p_j \, e^{-v_j(t)} }.
  \label{eq:h_ave}
\end{equation}
%
%
%
For the random noise part, we have
$\langle \zeta_i(t) \rangle = 0$,
%
%The noise is not necessarily white,
and the autocorrelation functions
depend only on the time difference
%
\begin{equation}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \sigma_{ij}(t - t'),
  \notag
  %\label{eq:zeta_zeta_correlation}
\end{equation}
%
where $\sigma_{ij}(t)$ is an even function
that vanishes at large $t$.
%
\note{More explicitly,
if the transition matrix of a period $\tau$
is $T^\tau_{ij}$,
then
$
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \pi_j - \pi_i \, \pi_j.
$
}%

@@

The bias potential, $v_i(t)$, following Eq. \eqref{eq:vt_diffeq}
generally grows overtime.
%
While a uniform increase of $v_i(t)$ does not affect the 
resulting distribution $\pi_i(t)$ as given by Eq. \eqref{eq:h_ave},
%
it is convenient to express the error of $v_i(t)$,
from the difference between $v_i(t)$ and a baseline value, $\bar v(t)$,
that represents the average growth\cite{
dama2014},
%
\begin{equation}
  x_i(t) \equiv v_i(t) - \bar v(t)
  %=
  %v_i - \sum_{j = 1}^n \rho_j \, v_j
  ,
\label{eq:x_def}
\end{equation}
%
where
%
\begin{equation}
  \bar v(t)
  = \sum_{i = 1}^n \rho_i \, v_i(t)
  ,
\label{eq:vbar_def}
\end{equation}
for some distribution,
$\pmb \rho = (\rho_1, \dots, \rho_n)$.
%
Apparently,
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, x_i(t) = 0
  .
\label{eq:rhox_sum}
\end{equation}
%
This distribution $\pmb \rho$
will be used to measure the error,
and is independent of
the target distribution $\mathbf p$.
%
For example,
in well-tempered metadynamics,
$\pmb\rho$ is a flat distribution,
while $\mathbf p$ is an attenuated version of
$\mathbf p^*$\cite{
barducci2008, dama2014}.
%
The rate of growth of $v(t)$ is given by
\begin{equation}
  \frac{ d \bar v (t) }
       { d t }
  =
  \alpha(t)
  \sum_{ r = 1 }^n \rho_r \, \frac{ h_r(t) } { p_r }
  .
\label{eq:dvbardt}
\end{equation}
%

In the asymptotic regime,
we expect the deviation $x_i(t)$
to be small.
%
Expanding Eq. \eqref{eq:h_ave} in $x_i$ yields
%
\begin{align}
  \frac{
    \langle h_i(t) \rangle
  }
  {
    p_i
  }
  &\approx
  \frac{                        e^{- x_i} }
       { \sum_{ j = 1}^n p_j \, e^{- x_j} }
  \approx
  1 - x_i + \sum_{r = 1}^n p_r \, x_r
  ,
\label{eq:hdet_linearize}
\end{align}
%
where we have used $\sum_{j=1}^n p_j = 1$.
%
\note{
The second step of Eq. \eqref{eq:hdet_linearize} follows
$$
\frac{ \langle h_i(t) \rangle }
     { p_i }
\approx
\frac{                       1 - x_i  }
     { \sum_{ r = 1 }^n p_r (1 - x_r) }
=
\frac{                       1 - x_i  }
     { 1 - \sum_{ r = 1 }^n p_r \, x_r }
.
$$
}%
From Eqs.
\eqref{eq:vt_diffeq},
\eqref{eq:h_split},
\eqref{eq:x_def}-\eqref{eq:hdet_linearize},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \xi_i(t) \right]
  ,
\label{eq:dxdt_WL}
\end{equation}
where
\begin{equation}
  \xi_i(t)
  =
  \frac{ \zeta_i(t) }
       { p_i }
  -
  \sum_{ r = 1 }^n
    \rho_r
    \frac{ \zeta_r(t) }
         {     p_r    }
  .
\label{eq:xi_def}
\end{equation}
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:dvbardt}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ h_i } { p_i }
    -
    \sum_{ r = 1 }^n
      \rho_r \frac{ h_r } { p_r }
  \right)
  \\
  &\stackrel{\mathrm{Eq.\; \eqref{eq:h_split}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ \langle h_i \rangle } { p_i }
    -
    \sum_{ r = 1 }^n
      \rho_r \frac{ \langle h_r \rangle } { p_r }
    +
    \xi_i(t)
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:rhox_sum} \;and\; \eqref{eq:hdet_linearize}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \xi_i(t) \right].
\end{aligned}
$$
}

We define
the total error of the bias potential as
%
\begin{equation}
  \Err
  =
  \sum_{i = 1}^n \rho_i \, \langle x_i^2 \rangle
  =
  \sum_{i = 1}^n \rho_i \, \left\langle (v_i - \bar v)^2 \right\rangle
  .
\label{eq:error_sum}
\end{equation}
%
Our aim is to find the $\alpha(t)$
that minimizes the error $E$
at the end of the simulation period $T$.
%
It suffices to consider
a simple one-variable problem.

@@

Consider the following equation
of a single variable:
%
\begin{equation}
\dot x(t) = -\alpha(t) \, \left[ x(t) - \xi(t) \right],
\label{eq:dxdt_alpha}
\end{equation}
%
where $\xi(t)$ is a generalized noise
that is invariant under time translation:
%
$
  \left\langle \xi(t) \, \xi(t') \right\rangle
  =
  \kappa(t - t')
  .
$
%
Additionally,
we assume that $\kappa(t)$ is an even function and
\begin{equation}
  \lim_{t \rightarrow \infty} \kappa(t)
  = 0
  .
  \label{eq:kappa_limit}
\end{equation}
%
For example, for a white noise,
$\kappa(t)$ is proportional to a
Dirac $\delta$-function, $\delta(t)$.
%
We wish to show that the $\alpha(t)$
of minimizing $\langle x^2(t) \rangle$ at long times
is given by
%
\begin{equation}
  \alpha(t) = \frac{ 1 }{ t + t_0 }.
\label{eq:alpha_invt1}
\end{equation}

@@

For the problem of
minimizing the error defined in Eq. \eqref{eq:error_sum}
for the single-bin scheme,
we note that Eq. \eqref{eq:dxdt_WL}
is of the form of Eq. \eqref{eq:dxdt_alpha}.
%
Thus,
the optimal schedule $\alpha(t)$ is given by
Eq. \eqref{eq:alpha_invt1}
for it optimizes each term $\langle x_i^2 \rangle$,
and hence the sum;
Eq. \eqref{eq:alpha_invt}
is the special case of $t_0 = 0$.
%
In Sec. \ref{sec:finlencorr},
we show that $t_0$ can be adjusted
to accommodate the initial error.

@@

The optimization procedure is similar to the single-bin case.
However, we shall adopt some restrictions and approximations
to simplify the discussion.

@@

The matrix $\mathbf w$ is not arbitrary.
%
First, to sample
the desired distribution
$\mathbf p = (p_1, \dots, p_n)$,
 $\dot v_i(t)$
in Eq. \eqref{eq:vt_diffeq_mbin}
should be a constant independent of $i$,
if $h_j(t)$ were the same as $p_j$.
%
This allows $v_i(t)$'s to grow uniformly
in the asymptotic regime.
%
Thus, by a proper scaling of $\alpha(t)$,
we may write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1
  .
\label{eq:w_sumj}
\end{equation}
%
In other words, $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
except that the elements can be negative.

@@

\item
Run a preliminary adaptive FDS simulation
%using the updating scheme
with a constant updating magnitude, $\alpha_0$.

>>>>

The Wang-Landau (WL) algorithm\cite{wang2001, wang2001pre}
and metadynamics\cite{huber1994, laio2002,
laio2008, barducci2011, sutto2012}
are two widely-used and closely-related\cite{micheletti2004}
flat-distribution-sampling (FDS) techniques
that build the bias potential adaptively or on the fly.
%
Both techniques employ regular updating of the bias potential
to discourage future visits to the current configuration
by incrementally elevating the bias potential there.
%
The difference lies
in the updating hill function,
which takes the form of a discrete 
$\delta$-function (or a boxcar function one histogram bin wide)
in the WL case,
or that of a Gaussian
in the metadynamics.
%
%is that in the WL case,
%each updating step is limited to the bin
%containing the current $z$,
%while metadynamics adopts an extended
%Gaussian window
%covering several neighboring bins.
%%
%The latter is more often used in MD simulations
%as it guarantees a smooth profile
%of the bias potential
%suitable for numerical differentiation
%in deriving the bias force.
%
In both cases,
regular updates to the bias potential
%in both cases, the bias potential
%is updated frequently (usually every few MC or MD steps),
%and the fluctuation
disrupts the underlying equilibrium
sampling\cite{laio2002},
and for convergence
one  has to gradually decrease
the updating magnitude.

The schedule of reducing
the updating magnitude,
therefore, critically affects
the precision of the final bias potential,
hence that of the PMF\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008, morozov2009,
komura2012, caparica2012, caparica2014,
barducci2008, dickson2011, dama2014}.
%
For the WL algorithm, it is known
that the optimal schedule
is a function of the
inverse time\cite{liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
morozov2007, zhou2008}
or
the inverse number of simulation steps.
%
The optimal schedule for the metadynamics case,
however, is less studied.

In this study,
we present a method of computing
the optimal schedule
for a general adaptive FDS method,
including the WL algorithm and metadynamics.
%
We will show that the optimal schedule can be derived
from an equation of motion of a free particle,
and the initial updating magnitude
should be set to
half of the previous equilibrium value.
%
Particularly, the optimal schedule in the WL case
is given by
an inverse-time formula,
with a modified origin of time.
%
However,
the optimal schedule for metadynamics is more complex
and sensitive to the simulation length
and the width of the Gaussian window.
%
Further, we generalize
the WL updating scheme
to a class of bandpass updating schemes
that are optimal for long simulations,
and the above origin-modified inverse-time schedule
turns out to be optimal for these schemes.
%
%For finite-length simulations, however,
%metadynamics may deliver better performance.
%
These results may help to better
plan and optimize FDS simulations.

@@

Consider the problem of computing
the distribution, $p^*_i$,
along a discrete quantity, $i$,
for a given system.
%
%The discreteness of $i$ is natural
%in discrete models.
%
For example, $i$ can be the energy, $E$,
in a lattice spin model or the temperature index
in a simulated tempering simulation\cite{
marinari1992, lyubartsev1992}.
%
For a continuous quantity, $z$,
which is often the case in a (classical) molecular system,
we can discretize $z$
such that the integer $i$ represents
the index of a small interval, or a bin,
$(z, z + \Delta z)$.
%
In both cases,
the distribution is normalized as
$\sum_{i = 1}^n p^*_i = 1$.

@@

Here we shall call a simulation that targets
a nearly flat distribution\cite{
dayal2004, trebst2004, zhang2007, barducci2008, singh2011},
$p_i$ constant for all $i$'s,
an FDS simulation.
%
Roughly speaking, there are two types of FDS methods.
%
The first type,
referred to as the equilibrium FDS, % (E-FDS) methods hereinafter,
uses a fixed bias potential, $V_i$,
(which is estimated beforehand)
during simulation.
%
Examples include the original
entropic or multicanonical sampling\cite{
berg1992, lee1993}
and simulated tempering\cite{
marinari1992, lyubartsev1992}
with $i$ being the index of energy
and that of temperature, respectively.
%
The estimated bias potentials at bin $i$, $V_i$,
are often improved
with the histogram of the sampling, $H_i$, accumulated
during simulation.
%
Using $H_i$ as an estimate of $\pi_i$ in
Eq. \eqref{eq:pi_p_phi1}
allows us to correct $V_i$ as
%
\begin{equation}
V^\mathrm{corr}_i
=
V_i
+
\ln \frac{ H_i }
         { p_i }.
\label{eq:vcorr_equil}
\end{equation}
%
\note{This follows from
  $$
  \begin{aligned}
    H_i \approx \pi_i
    &\propto p^*_i \, e^{-V_i},
    \\
    p_i
    &\propto p^*_i \, e^{-V^\mathrm{corr}_i}.
  \end{aligned}
  $$
}
For an unknown system,
it is often difficult
to guess a sufficiently accurate
$V_i$ with which to start.
%
Thus,
it can take many iterative rounds of simulations
to achieve the desired flat distribution.
%
\note{Further, a reasonably precise histogram
requires an accumulation period
longer than the autocorrelation time,
which can also be hard to estimate beforehand.}

For this reason, the second type of methods,
referred to as the adaptive FDS methods,
were invented.
%
Examples include
the WL algorithm and metadynamics.
%
These methods incrementally update the bias potential
and can approximately sample the target distribution.
%
However,
%they are less well understood theoretically, as
as mentioned, the on-the-fly updates break the microscopic reversibility
of the underlying sampling mechanism.
Further, failure to properly reduce the size of the histogram increments
would lead to roughness or errors in the final estimate.
%
%Thus, for a conservative practitioner,
%the E-FDS methods appear more appealing
%once a sufficiently accurate bias potential is available.
%
To minimize this effect,
many adaptive methods adopt a schedule to
reduce the updating magnitude
over the course of simulation\cite{
marsili2006,
liang2007,
belardinelli2007, belardinelli2007jcp, belardinelli2008,
barducci2008}.
%
Thus, in late stages of a long simulation,
the updating becomes negligible and
%the adaptive simulation tends to
%the corresponding equilibrium one.
%
if the bias potential were converged to the exact value,
this adaptive FDS tends to the ideal equilibrium counterpart,
%particular equilibrium simulation can be thought as
%the optimal asymptotic limit of the adaptive counterpart,
delivering the maximal efficiency.

@@

One can show that with a constant magnitude $\alpha(t) = \alpha > 0$,
the distribution collected from
an adaptive FDS simulation
is identical to $p_i$ in long times.
%
However, as the above updates
change $v_i(t)$ continuously,
Eq. \eqref{eq:pi_p_v} no longer holds
at all times,
and the error in $v_i(t)$
depends on the updating magnitude\cite{
  zhou2005, liang2005, laio2005, bussi2006, poulain2006, liang2007,
  morozov2007, zhou2008, morozov2009, crespo2010,
  atchade2011, fort2015}.
%%
%The source of the deviation is two-fold.
%%
%First, since $v_i(t)$ is updated continuously,
%there is a random noise that is proportional
%to $\sqrt \alpha$.
%%
%Second, there is a systematic error
%that comes from the updating dynamics itself,
%as it breaks the Markovian nature
%that underlies Eq. \eqref{eq:pi_p_v}.
%%
%
Thus, one has to reduce the magnitude $\alpha$ over time
to improve the estimate of the distribution,
$p^*_i$. %via Eq. \eqref{eq:v_def}.

@@

We first approximate the finite difference Eq. \eqref{eq:wl_update}
by a differential equation
%
\begin{equation}
  \dot v_i(t)
  \equiv
  \frac{ dv_i(t) } { dt }
  =
  \alpha(t) \, \frac{ h_i(t) } { p_i },
  \label{eq:vt_diffeq}
\end{equation}
%
with
%
\begin{equation}
  h_i(t) = \delta_{i, i(t)}
  ,
  \label{eq:h_def}
\end{equation}
%
being the instantaneous histogram,
which is equal to $1.0$
for the bin $i(t)$ at time $t$
or zero otherwise.

Next, we split $h_i(t)$ into a deterministic averaged part
and a stochastic part, $\zeta_i(t)$,
%
\begin{equation}
  h_i(t) = \langle h_i(t) \rangle + \zeta_i(t).
  \label{eq:h_split}
\end{equation}
%
The deterministic part can be related
to the ``ensemble average'' of $h_i$,
where the ensemble consists of many similar simulation copies
sharing the same schedule $\alpha(t)$
the same $v_i(t)$ at time $t$.
%
The initial states and the stochastic forces
during the process may differ, however.

For sufficiently small $\alpha(t)$,
the $v_i$'s remain roughly the same for a short period.
%
Then,
the sampling process is approximately Markovian, % of a finite order,
and we may assume Eq. \eqref{eq:pi_p_v}
for the deterministic part
%
\begin{equation}
  \langle h_i(t) \rangle
  \approx
  \pi_i(t)
  =
  \frac{                p_i \, e^{-v_i(t)} }
       { \sum_{j = 1}^n p_j \, e^{-v_j(t)} }.
  \label{eq:h_ave}
\end{equation}
%
%
%
For the random noise part,
the average is zero by definition,
$\langle \zeta_i(t) \rangle = 0$,
%
%The noise is not necessarily white,
and the autocorrelation functions
depend only on the time difference
%
\begin{equation}
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \sigma_{ij}(t - t'),
  \notag
  %\label{eq:zeta_zeta_correlation}
\end{equation}
%
where $\sigma_{ij}(t)$ is an even function
that vanishes at large $t$.
%
\note{More explicitly,
if the transition matrix of a period $\tau$
is $T^\tau_{ij}$,
then
$
  \langle \zeta_i(t) \, \zeta_j(t') \rangle
  =
  \langle h_i(t) \, h_j(t') \rangle
  -
  \langle h_i(t) \rangle \, \langle h_j(t') \rangle
  =
  T^{t - t'}_{ij} \pi_j - \pi_i \, \pi_j.
$
}%

@@

The bias potential, $v_i(t)$, following Eq. \eqref{eq:vt_diffeq}
generally grows overtime.
%
Since a uniform increase of $v_i(t)$ does not affect the
resulting distribution $\pi_i(t)$ as given by Eq. \eqref{eq:h_ave},
%
the real error in $v_i(t)$ is more readily defined
from the difference between $v_i(t)$ and a baseline value, $\bar v(t)$,
that represents the average growth\cite{
dama2014},
%
\begin{equation}
  x_i(t) \equiv v_i(t) - \bar v(t)
  %=
  %v_i - \sum_{j = 1}^n \rho_j \, v_j
  ,
\label{eq:x_def}
\end{equation}
%
where
%
\begin{equation}
  \bar v(t)
  = \sum_{i = 1}^n \rho_i \, v_i(t)
  ,
\label{eq:vbar_def}
\end{equation}
for some distribution,
$\pmb \rho = (\rho_1, \dots, \rho_n)$.
%
So we have,
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, x_i(t) = 0
  .
\label{eq:rhox_sum}
\end{equation}
%
This distribution $\pmb \rho$
will be used to measure the error,
and is independent of
the target distribution $\mathbf p$.
%
For example,
in well-tempered metadynamics,
$\pmb\rho$ is a flat distribution,
while $\mathbf p$ is an attenuated version of
$\mathbf p^*$\cite{
barducci2008, dama2014}.
%
The rate of growth of $v(t)$ is given by
\begin{equation}
  \frac{ d \bar v (t) }
       { d t }
  =
  \alpha(t)
  \sum_{ r = 1 }^n \rho_r \, \frac{ h_r(t) } { p_r }
  .
\label{eq:dvbardt}
\end{equation}
%

In the asymptotic regime,
we expect the deviation $x_i(t)$
to be small.
%
Expanding Eq. \eqref{eq:h_ave} in $x_i$ yields
%
\begin{align}
  \frac{
    \langle h_i(t) \rangle
  }
  {
    p_i
  }
  &\approx
  \frac{                        e^{- x_i} }
       { \sum_{ j = 1}^n p_j \, e^{- x_j} }
  \approx
  1 - x_i + \sum_{r = 1}^n p_r \, x_r
  ,
\label{eq:hdet_linearize}
\end{align}
%
where we have used $\sum_{j=1}^n p_j = 1$.
%
\note{
The second step of Eq. \eqref{eq:hdet_linearize} follows from
$$
\frac{ \langle h_i(t) \rangle }
     { p_i }
\approx
\frac{                       1 - x_i  }
     { \sum_{ r = 1 }^n p_r (1 - x_r) }
=
\frac{                       1 - x_i  }
     { 1 - \sum_{ r = 1 }^n p_r \, x_r }
.
$$
}%
From Eqs.
\eqref{eq:vt_diffeq},
\eqref{eq:h_split},
\eqref{eq:x_def}-\eqref{eq:hdet_linearize},
we get a set of decoupled equations
%
\begin{equation}
  \dot x_i(t)
  =
  -\alpha(t) \, \left[ x_i(t) - \xi_i(t) \right]
  ,
\label{eq:dxdt_WL}
\end{equation}
where
\begin{equation}
  \xi_i(t)
  =
  \frac{ \zeta_i(t) }
       { p_i }
  -
  \sum_{ r = 1 }^n
    \rho_r
    \frac{ \zeta_r(t) }
         {     p_r    }
  .
\label{eq:xi_def}
\end{equation}
\note{The derivation:
$$
\begin{aligned}
  \dot x_i
  &\stackrel{\mathrm{Eq.\; \eqref{eq:x_def}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \dot v_i - \dot{ \bar v }
  \stackrel{\mathrm{Eqs.\; \eqref{eq:vt_diffeq}\; and\; \eqref{eq:dvbardt}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ h_i } { p_i }
    -
    \sum_{ r = 1 }^n
      \rho_r \frac{ h_r } { p_r }
  \right)
  \\
  &\stackrel{\mathrm{Eq.\; \eqref{eq:h_split}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left(
    \frac{ \langle h_i \rangle } { p_i }
    -
    \sum_{ r = 1 }^n
      \rho_r \frac{ \langle h_r \rangle } { p_r }
    +
    \xi_i(t)
  \right)
  \\
  &\stackrel{\mathrm{Eqs.\; \eqref{eq:rhox_sum} \;and\; \eqref{eq:hdet_linearize}}}
            {=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=\joinrel=}
  \alpha(t) \left[ -x_i(t) + \xi_i(t) \right].
\end{aligned}
$$
}

We define
the total error of the bias potential as
%
\begin{equation}
  \Err
  =
  \sum_{i = 1}^n \rho_i \, \langle x_i^2 \rangle
  =
  \sum_{i = 1}^n \rho_i \, \left\langle (v_i - \bar v)^2 \right\rangle
  .
\label{eq:error_sum}
\end{equation}
%
The expression depends implicitly on the function $\alpha(t)$.
%
Our aim is to find the $\alpha(t)$
that minimizes the error $E$
at the end of the simulation period $T$.
%
It suffices to consider
a one-variable problem that
minimizes each individual $\left\langle x_i^2 \right\rangle$.

@@

Consider the following stochastic differential equation
of a single variable:
%
\begin{equation}
\dot x(t) = -\alpha(t) \, \left[ x(t) - \xi(t) \right],
\label{eq:dxdt_alpha}
\end{equation}
%
where $\xi(t)$ is a generalized noise
that is invariant under time translation:
%
$
  \left\langle \xi(t) \, \xi(t') \right\rangle
  =
  \kappa(t - t')
  .
$
%
Additionally,
we assume that $\kappa(t)$ is an even function and
\begin{equation}
  \lim_{t \rightarrow \infty} \kappa(t)
  = 0
  .
  \label{eq:kappa_limit}
\end{equation}
%
For example, for a white noise,
$\kappa(t)$ is proportional to a
Dirac $\delta$-function, $\delta(t)$.
%
We wish to show that the $\alpha(t)$
of minimizing $\langle x^2(t) \rangle$ at long times
is given by
%
\begin{equation}
  \alpha(t) = \frac{ 1 }{ t + t_0 }.
\label{eq:alpha_invt1}
\end{equation}

@@

For the problem of
minimizing the error defined in Eq. \eqref{eq:error_sum}
for the single-bin scheme,
we note that Eq. \eqref{eq:dxdt_WL}
is of the form of Eq. \eqref{eq:dxdt_alpha}.
%
Thus,
the optimal schedule $\alpha(t)$ is given by
Eq. \eqref{eq:alpha_invt1}
for it optimizes each term $\langle x_i^2 \rangle$,
and hence the sum;
Eq. \eqref{eq:alpha_invt}
is the special case of $t_0 = 0$.
%
In Sec. \ref{sec:finlencorr},
we show that % $t_0$ can be adjusted
one can further improve the schedule by adjusting $t_0$
to accommodate the initial error.

@@

The optimization procedure is similar to the single-bin case.
%
However, we shall assume some restrictions and approximations
to simplify the discussion.

@@

The matrix $\mathbf w$ is not arbitrary.
%
First, to sample
the desired distribution
$\mathbf p = (p_1, \dots, p_n)$,
 $\dot v_i(t)$
in Eq. \eqref{eq:vt_diffeq_mbin}
should be a constant of $i$,
if $h_j(t)$ were the same as $p_j$.
%
This allows $v_i(t)$'s to grow uniformly
in the asymptotic regime.
%
Thus, by a proper scaling of $\alpha(t)$,
we may write this condition as
%
\begin{equation}
  \sum_{j = 1}^n w_{ij} = 1
  .
\label{eq:w_sumj}
\end{equation}
%
In other words, $(1, \dots, 1)^T$
is a right eigenvector of $\mathbf w$
with eigenvalue $1$.
%
Thus, the transpose $\mathbf w^T$
resembles a transition matrix,
except that some elements can be negative.
%
Below we shall take advantage of the analogy
and borrow techniques used
in studying transition matrices\cite{vankampen}.

@@

\item
Run a preliminary adaptive FDS simulation
%using the updating scheme
with a constant updating magnitude, $a_0$.
